---
title: "Labour Market in Colombia"
subtitle: "Indices"
author: "Miguel Martinez"
date: today
format: 
  html:
    toc: true
    theme: cosmo
editor: visual
execute:
  warning: false
---

This document presents the process of building some graphs of the evolution of the unemployment, employment and labor force participation rate in Colombia for the last years. The data is made public by the Departamento Administrativo de Colombia ([DANE](https://www.dane.gov.co/)), and you can reproduce this downloading the data set [here](https://www.dane.gov.co/index.php/estadisticas-por-tema/mercado-laboral/empleo-y-desempleo).

The first step is to load the following libraries:

```{r}

library(readxl)
library(tidyverse)
```

Now we can load the datasets:

## Load data

```{r}

# Here we load the labour market indices for Total country

df1 = readxl::read_excel("anex-GEIH-Desestacionalizado-abr2025.xlsx",
                         sheet = "Total nacional",
                         range = "A12:KG16",
                         col_names = FALSE)

# Now we load the labour market indices for Total 13 cities and metropolitan areas

df2 = readxl::read_excel("anex-GEIH-Desestacionalizado-abr2025.xlsx",
                         sheet = "Total 13 ciudades A.M.",
                         range = "A12:KG16",
                         col_names = FALSE)
```

## Indexing

```{r}

# Indexing the year and month variables to the dataframes

df1[1, 1] = "Año"
df1[2, 1] = "Mes"

df2[1, 1] = "Año"
df2[2, 1] = "Mes"
```

## Lengthening Data

```{r}

# Total country

df1 = df1 |>
  tidyr::pivot_longer(cols = 2:293,
                      values_to = "value") |> 
  tidyr::fill(value) |> 
  tidyr::pivot_wider(names_from = 1,
                     values_from = value) |> 
  dplyr::select(Año:`Tasa de Desocupación (TD)`)

```

```{r}

# Total 13 cities and metropolitan areas

df2 = df2 |>
  tidyr::pivot_longer(cols = 2:293,
                      values_to = "value") |> 
  tidyr::fill(value) |> 
  tidyr::pivot_wider(names_from = 1,
                     values_from = value) |> 
  dplyr::select(Año:`Tasa de Desocupación (TD)`)

```

## Merging Dataframes

```{r}

# Now we merge both dataframes: total country and 13 cities and metropolitan areas labour market indices by (año and mes)

data = df1 |> 
  dplyr::inner_join(df2,
                    by = c("Año", "Mes"),
                    suffix = c(".TN", ".13 A.M"))

data
```

## D type transformation

```{r}

# Now we transform some variables to numeric

data = data |> 
  dplyr::mutate_at(vars(1, 3:8), as.numeric)

glimpse(data)
```

```{r}

# Please first notice that some months have asterisks, so we  need to replace them too

meses = data |>
  dplyr::distinct(Mes)

meses$Mes
```

```{r}

# Now we transform the "string" format of months to "numeric"

data = data |>
  dplyr::mutate(
    Mes = recode(Mes,
                 "Ene" = 01, "Feb" = 02,
                 "Mar" = 03, "Abr" = 04,
                 "May" = 05, "Jun" = 06,
                 "Jul" = 07, "Ago" = 08,
                 "Sep" = 09, "Oct" = 10,
                 "Nov" = 11, "Dic" = 12,
                 "Jul*" = 07, "Ago*" = 08))

glimpse(data)
```

## Filtering

```{r}

# Then, we filter the data set from 2019:

data = data |>
  dplyr::filter(Año >= 2019)

```

## Unemployment

Now we filter the data for unemployment and then present the data in the *long* or *stacked* format:

```{r}

unemp = data |>
  dplyr::select(Año, Mes,
                `Tasa de Desocupación (TD).TN`,
                `Tasa de Desocupación (TD).13 A.M`) |>
  tidyr::pivot_longer(cols = 3:4,
                      names_to = "TD",
                      values_to = "value")  

unemp
```

```{r}

unemp |> 
  ggplot(aes(x = Mes, y = value, colour = TD, group = TD)) +
  geom_point() +
  geom_line() +
  facet_grid( . ~ Año) +
  scale_x_continuous(breaks = seq(0,12, by=2)) +
  scale_y_continuous(breaks = seq(8,26, by=2)) +
  theme_classic() +
  labs(title = "Tasa de desocupación – TD desestacionalizada (preliminar).
       Total nacional y 13 ciudades y áreas metropolitanas",
       subtitle = "Mensual. enero 2019 – abril 2025",
       x = "Mes",
       y = "Tasa de Desocupación (%)",
       caption = "Fuente: Elaboración propia con datos del DANE, GEIH") +
  theme(plot.caption.position = "panel",
        plot.caption = element_text(hjust = 0),
        legend.position = "bottom")

```

## Employment

Now we filter the data for employment and then present the data in the *long* or *stacked* format:

```{r}

emp = data |>
  dplyr::select(Año, Mes,
                `Tasa de Ocupación (TO).TN`,
                `Tasa de Ocupación (TO).13 A.M`) |>
  tidyr::pivot_longer(cols = 3:4,
                      names_to = "TO",
                      values_to = "value")  

emp
```

```{r}

emp |> 
  ggplot(aes(x = Mes, y = value, colour = TO, group = TO)) +
  geom_point() +
  geom_line() +
  facet_grid( . ~ Año) +
  scale_x_continuous(breaks = seq(0,12, by=2)) +
  theme_classic() +
  labs(title = "Tasa de ocupación – TO desestacionalizada (preliminar).
       Total nacional y 13 ciudades y áreas metropolitanas",
       subtitle = "Mensual. enero 2019 – abril 2025",
       x = "Mes",
       y = "Tasa de Ocupación (%)",
       caption = "Fuente: Elaboración propia con datos del DANE, GEIH") +
  theme(plot.caption.position = "panel",
        plot.caption = element_text(hjust = 0),
        legend.position = "bottom") 

```

## Labor Force Participation

Now we filter the data for labor force participation rate and then present the data in the *long* or *stacked* format:

```{r}

lfp = data |>
  dplyr::select(Año, Mes,
                `Tasa Global de Participación (TGP).TN`,
                `Tasa Global de Participación (TGP).13 A.M` ) |>
  tidyr::pivot_longer(cols = 3:4,
                      names_to = "TGP",
                      values_to = "value")

lfp
```

```{r}

lfp |> 
  ggplot(aes(x = Mes, y = value, colour = TGP, group = TGP)) +
  geom_point() +
  geom_line() +
  facet_grid( . ~ Año) +
  scale_x_continuous(breaks = seq(0,12, by=2)) +
  theme_classic() +
  labs(title = "Tasa global de participación – TGP desestacionalizada (preliminar).
       Total nacional y 13 ciudades y áreas metropolitanas",
       subtitle = "Mensual. enero 2019 – abril 2025",
       x = "Mes",
       y = "Tasa global de participación (%)",
       caption = "Fuente: Elaboración propia con datos del DANE, GEIH") +
  theme(plot.caption.position = "panel",
        plot.caption = element_text(hjust = 0),
        legend.position = "bottom") 

```

## Conclusion

First, we can notice that the main labor market rates are highly associated with the economic activity. You can observe that the unemployment rate has stabilized at levels slightly below the pre-pandemic numbers, but at degrees still considered highly for this country. On the other hand, the labor market is also observing growth in the employment rates. The analysts and policy-makers seek to achieve an economic growth of 2.6% in 2025. The Colombian economy is facing a strong challenge in the past years in reducing the unemployment rate in the long run, mainly because there are divergences in the jobs creation of the formal relative to the informal sector, and because of the excessive burdens the companies face at hiring people, summed to the economic policies of the minimum wage that are insufficient to reach the most labor market people.
