---
title: "Mercado Laboral en Colombia"
subtitle: "Tasa de desempleo"
author: "Miguel Martinez"
date: today
format: 
  html:
    toc: true
    theme: cosmo
editor: visual
execute:
  warning: false
---

This document presents the process of building a graph of the evolution of the unemployment rate in Colombia. The data is made public by the Departamento Administrativo de Colombia ([DANE](https://www.dane.gov.co/)), and you can reproduce this downloading the data set [here](https://www.dane.gov.co/index.php/estadisticas-por-tema/mercado-laboral/empleo-y-desempleo).

The first step is to load the following libraries:

```{r}

library(readxl)
library(tidyverse)
```

Now we can load the data set:

## Load data

```{r}

data = read_excel("data.xlsx")

data
```

Then, we filter the data set from 2019:

```{r}

data = data |> 
  filter(Año>=2019)

data
```

## Unemployment

Now we filter the data for unemployment and then present the data in the *long* or *stacked* format:

```{r}

unemp = data |> 
  select(Año, Mes, `Tasa de Desocupación (TD) TN`, `Tasa de Desocupación (TD) Total 13 ciudades A.M.`) |> 
  pivot_longer(cols = 3:4, names_to = "TD",
               values_to = "value")

unemp
```

```{r}

ggplot(unemp) +
  geom_point(aes(x = Mes, y = value, colour = TD, group = TD)) +
  geom_line(aes(x = Mes, y = value, colour = TD, group = TD)) +
  facet_grid( . ~ Año) +
  scale_x_continuous(breaks = seq(0,12, by=2)) +
  scale_y_continuous(breaks = seq(8,26, by=2)) +
  theme_classic() +
  labs(title = "Tasa de desocupación – TD desestacionalizada (preliminar).
       Total nacional y 13 ciudades y áreas metropolitanas",
       subtitle = "Mensual. enero 2019 – marzo 2025",
       x = "Mes",
       y = "Tasa de Desocupación (%)",
       caption = "Fuente: Elaboración propia con datos del DANE, GEIH") +
  theme(plot.caption.position = "panel",
        plot.caption = element_text(hjust = 0),
        legend.position = "bottom")

```

## Employment

Now we filter the data for employment and then present the data in the *long* or *stacked* format:

```{r}

emp = data |> 
  select(Año, Mes, `Tasa de Ocupación (TO) TN`, `Tasa de Ocupación (TO) Total 13 ciudades A.M.`) |> 
  pivot_longer(cols = 3:4, names_to = "TO",
               values_to = "value")

emp
```

```{r}

ggplot(emp) +
  geom_point(aes(x = Mes, y = value, colour = TO, group = TO)) +
  geom_line(aes(x = Mes, y = value, colour = TO, group = TO)) +
  facet_grid( . ~ Año) +
  scale_x_continuous(breaks = seq(0,12, by=2)) +
  theme_classic() +
  labs(title = "Tasa de ocupación – TO desestacionalizada (preliminar).
       Total nacional y 13 ciudades y áreas metropolitanas",
       subtitle = "Mensual. enero 2019 – marzo 2025",
       x = "Mes",
       y = "Tasa de Ocupación (%)",
       caption = "Fuente: Elaboración propia con datos del DANE, GEIH") +
  theme(plot.caption.position = "panel",
        plot.caption = element_text(hjust = 0),
        legend.position = "bottom")

```

## Labor Force Participation

Now we filter the data for labor force participation rate and then present the data in the *long* or *stacked* format:

```{r}

lfp = data |> 
  select(Año, Mes, `Tasa Global de Participación (TGP) TN`, `Tasa Global de Participación (TGP) Total 13 ciudades A.M.`) |> 
  pivot_longer(cols = 3:4, names_to = "TGP",
               values_to = "value")

lfp
```

```{r}

ggplot(lfp) +
  geom_point(aes(x = Mes, y = value, colour = TGP, group = TGP)) +
  geom_line(aes(x = Mes, y = value, colour = TGP, group = TGP)) +
  facet_grid( . ~ Año) +
  scale_x_continuous(breaks = seq(0,12, by=2)) +
  theme_classic() +
  labs(title = "Tasa global de participación – TGP desestacionalizada (preliminar).
       Total nacional y 13 ciudades y áreas metropolitanas",
       subtitle = "Mensual. enero 2019 – marzo 2025",
       x = "Mes",
       y = "Tasa global de participación (%)",
       caption = "Fuente: Elaboración propia con datos del DANE, GEIH") +
  theme(plot.caption.position = "panel",
        plot.caption = element_text(hjust = 0),
        legend.position = "bottom")

```

## Conclusion

First, we can notice that the main labor market rates are highly associated with the economic activity. You can observe that the unemployment rate has stabilized at levels slightly below the pre-pandemic numbers, but at degrees still considered highly for this country. On the other hand, the labor market is also observing growth in the employment rates. The analysts and policy-makers seek to achieve an economic growth of 2.6% in 2025. The Colombian economy is facing a strong challenge in the past years in reducing the unemployment rate in the long run, mainly because there are divergences in the jobs creation of the formal relative to the informal sector, and because of the excessive burdens the companies face at hiring people, summed to the economic policies of the minimum wage that are insufficient to reach the most labor market people.
